---
title: "re2_getting creative with ozone concentration"
author: "keerthi"
date: "2025-03-10"
output:
  pdf_document: default
  html_document: default
---

```{r r_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# installing packages in case not installed
use_pkgs <- c("dplyr", "rmarkdown", "tidyverse","knitr", "tidyr", "ggplot2", "lubridate", "corrplot", "car", "viridis")  
new_pkgs <- use_pkgs[!(use_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs) > 0) install.packages(new_pkgs,repos = "http://cran.us.r-project.org")
invisible(lapply(use_pkgs, require, character.only = TRUE))

# loading required libraries
library(dplyr)
library(rmarkdown)
library(tidyverse)
library(knitr)
library(here)
library(tidyr)
library(ggplot2)
library(lubridate)
library(corrplot)
library(car)
library(viridis)
```


```{r r_analysis1, echo=FALSE, message=FALSE, warning=FALSE}

# introduction: this report exercise shows how ozone concentration is influenced by other variables like solar radiation, wind and temperature. To be specific, regression of the ozone variable on other variables (e.g., temperature, solar radiation, wind) is estimated for a directional relationship. Besides, correlation is also estimated to measure the strength and direction of linear association between variables.

here::here("data/RE2_ozone_data.csv")

ozone <- read.csv("../data/RE2_ozone_data.csv")


# reshaping data from wide to long for easier plotting of multiple variables
ozone_long <- ozone %>%
  pivot_longer(cols = c(Ozone, Solar.R, Wind, Temp), names_to = "Variable", values_to = "Value")


# cleaning the data- removal of NA and imputing
# imputing missing values in solar.R
mean_solar <- mean(ozone$Solar.R, na.rm = TRUE)
ozone$Solar.R[is.na(ozone$Solar.R)] <- mean_solar

# remove rows where Ozone is NA
ozone_clean <- ozone%>% filter(!is.na(Ozone))

# only keep relevant numeric columns (removing rows with NA)
ozone_clean <- ozone %>%
  select(Ozone, Solar.R, Wind, Temp) %>%
  drop_na()


# boxplot- visualization 1

# plot boxplots for each variable grouped by Month
ggplot(ozone_long, aes(x = factor(Month), y = Value)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.25) +  # adds the horizontal bar (whisker caps)
  facet_wrap(~ Variable, scales = "free_y") +
  labs(x = "Month", y = "Value", title = "Monthly Boxplots of Variables") +
  theme_minimal()

ggplot(ozone_long, aes(x = factor(Month), y = Value)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_jitter(width = 0.2, alpha = 0.5) +  # add daily points with some horizontal jitter
  facet_wrap(~ Variable, scales = "free_y") +
  labs(x = "Month", y = "Value", title = "Monthly Boxplots with Daily Points") +
  theme_minimal()


# statistical metrics- mean, median, range

# mean
monthly_mean_values <- ozone %>%
  group_by(Month) %>%
  summarise(
    mean_Ozone = mean(Ozone, na.rm = TRUE),
    mean_Solar = mean(Solar.R, na.rm = TRUE),
    mean_Wind = mean(Wind, na.rm = TRUE),
    mean_Temp = mean(Temp, na.rm = TRUE)
  )
print(monthly_mean_values)

# median
monthly_median_values <- ozone %>%
  group_by(Month) %>%
  summarise(
    median_Ozone = median(Ozone, na.rm = TRUE),
    median_Solar = median(Solar.R, na.rm = TRUE),
    median_Wind = median(Wind, na.rm = TRUE),
    median_Temp = median(Temp, na.rm = TRUE)
  )

print(monthly_median_values)

# range 
monthly_range_values <- ozone %>%
  group_by(Month) %>%
  summarise(
    range_Ozone = max(Ozone, na.rm = TRUE) - min(Ozone, na.rm = TRUE),
    range_Solar = max(Solar.R, na.rm = TRUE) - min(Solar.R, na.rm = TRUE),
    range_Wind = max(Wind, na.rm = TRUE) - min(Wind, na.rm = TRUE),
    range_Temp = max(Temp, na.rm = TRUE) - min(Temp, na.rm = TRUE)
  )

print(monthly_range_values)

# max, min and range by month
monthly_max_min_range <- ozone %>%
  group_by(Month) %>%
  summarise(
    max_Ozone = max(Ozone, na.rm = TRUE),
    min_Ozone = min(Ozone, na.rm = TRUE),
    range_Ozone = max_Ozone - min_Ozone,
    
    max_Solar = max(Solar.R, na.rm = TRUE),
    min_Solar = min(Solar.R, na.rm = TRUE),
    range_Solar = max_Solar - min_Solar,
    
    max_Wind = max(Wind, na.rm = TRUE),
    min_Wind = min(Wind, na.rm = TRUE),
    range_Wind = max_Wind - min_Wind,
    
    max_Temp = max(Temp, na.rm = TRUE),
    min_Temp = min(Temp, na.rm = TRUE),
    range_Temp = max_Temp - min_Temp
  )

print(monthly_max_min_range)

# correlation matrix- visualization 2

# ozone vs solar.R, Wind, Temp
cor(ozone_clean$Ozone, ozone_clean$Solar.R)  # Ozone vs Solar.R
cor(ozone_clean$Ozone, ozone_clean$Wind)     # Ozone vs Wind
cor(ozone_clean$Ozone, ozone_clean$Temp)     # Ozone vs Temp

cor(ozone_clean)

# correlation matrix_visualization
corr_matrix <- cor(ozone_clean)
corrplot(corr_matrix, method = "circle", type = "lower", tl.cex = 0.8)


#regression analysis

model <- lm(Ozone ~ Solar.R + Wind + Temp, data = ozone_clean)
summary(model)

#checking multicollinearity
vif(model)


# regression with sub plots- visualization 3
ozone_raw <- ozone %>%
  mutate(Date = make_date(year = 1973, month = Month, day = Day))

ozone_clean_month <- ozone_raw
ozone_clean_month$month_name <- month(ozone_raw$Month, label = TRUE)

# ozone vs solar.r
ggplot(
  data = ozone_clean_month,
  aes(x = Solar.R, y = Ozone)
) +
  geom_point(alpha = 0.4) +
  geom_smooth(formula = y ~ x + 0, method = "lm", color = "red", se = FALSE) +
  labs(
    x = expression(paste("Solar.R (lang)")),
    y = expression(paste("Ozone (ppb)")),
    caption = "Scatter plot of Ozone vs Solar Radiation for each month from May to September. 
               Red lines show linear regression fits with zero intercept."
  ) +
  facet_wrap(~month_name) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, margin = margin(t = 10, b = 10), size = 9),
    plot.caption.position = "plot"
  )

# ozone vs wind
ggplot(
  data = ozone_clean_month,
  aes(x = Wind, y = Ozone)
) +
  geom_point(alpha = 0.4) +
  geom_smooth(formula = y ~ x + 0, method = "lm", color = "red", se = FALSE) +
  labs(
    x = expression(paste("Wind (mph)")),
    y = expression(paste("Ozone (ppb)")),
    caption = "Scatter plot of Ozone vs Wind for each month from May to September. 
               Red lines show linear regression fits with zero intercept."
  ) +
  facet_wrap(~month_name) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, margin = margin(t = 10, b = 10), size = 9),
    plot.caption.position = "plot"
  )

# ozone vs Temp
ggplot(
  data = ozone_clean_month,
  aes(x = Temp, y = Ozone)
) +
  geom_point(alpha = 0.4) +
  geom_smooth(formula = y ~ x + 0, method = "lm", color = "red", se = FALSE) +
  labs(
    x = expression(paste("Temp (°F)")),
    y = expression(paste("Ozone (ppb)")),
    caption = "Scatter plot of Ozone vs Temperature for each month from May to September. 
               Red lines show linear regression fits with zero intercept."
  ) +
  facet_wrap(~month_name) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, margin = margin(t = 10, b = 10), size = 9),
    plot.caption.position = "plot"
  )


# time series- temporal analysis: visualization 4

# adding date from raw data 
ozone_raw <- ozone %>%
  mutate(Date = make_date(year = 1973, month = Month, day = Day))

# Create a copy of ozone_clean
ozone_clean_new <- ozone_clean

# Add the Date column from ozone_raw, but only up to the number of rows in ozone_clean
ozone_clean_new$Date <- ozone_raw$Date[1:nrow(ozone_clean_new)]

# Ozone vs Solar.R
ggplot(data = ozone_clean_new, aes(x = Date, y = Ozone)) +
  geom_line() +
  geom_point(aes(color = Solar.R), size = 1) +
  labs(
    title = "Ozone Levels Over Time (colored by Solar.R)",
    x = "Date",
    y = "Ozone (ppb)",
    color = "Solar.R (lang)"
  ) +
  scale_color_viridis_c(option = "D", direction = -1) +
  theme_classic()

# Ozone vs Wind
ggplot(data = ozone_clean_new, aes(x = Date, y = Ozone)) +
  geom_line() +
  geom_point(aes(color = Wind), size = 1) +
  labs(
    title = "Ozone Levels Over Time (colored by Wind)",
    x = "Date",
    y = "Ozone (ppb)",
    color = "Wind (mph)"
  ) +
  scale_color_viridis_c(option = "D", direction = -1) +
  theme_classic()

# Ozone vs Temp
ggplot(data = ozone_clean_new, aes(x = Date, y = Ozone)) +
  geom_line() +
  geom_point(aes(color = Temp), size = 1) +
  labs(
    title = "Ozone Levels Over Time (colored by Temp)",
    x = "Date",
    y = "Ozone (ppb)",
    color = "Temp (°F)"
  ) +
  scale_color_viridis_c(option = "D", direction = -1) +
  theme_classic()


# tables from knitr function- Mean, Median, Range

knitr::kable(monthly_mean_values, caption = "Table 1: Monthly Mean of Ozone, Solar Radiation, Wind, and Temperature.")
knitr::kable(monthly_median_values, caption = "Table 2: Monthly Median of Ozone, Solar Radiation, Wind, and Temperature.")
knitr::kable(monthly_range_values, caption = "Table 3: Monthly Ranges of Ozone, Solar Radiation, Wind, and Temperature.")


```
# Dataset Description- This report analyzes diurnal air quality measurements from New York across three sites—Roosevelt Island, Central Park, and LaGuardia Airport—recorded between May and September 1973 (153 observations). Variables include mean ozone levels (ppb), solar radiation (Langleys), average wind speed (mph), and maximum daily temperature (°F).

# Data Distribution and Outliers- To understand variable distribution and detect outliers, monthly boxplots (visualisation 1) were created. Ozone showed right skewness with high-value outliers, except in July. Solar radiation displayed left skewness from July to September, with no outliers in August and September—indicating seasonal daylight decline. Median patterns in ozone and solar radiation, and later temperature, appeared aligned. Wind speeds had few outliers and showed a slight rise in September, consistent with early autumn. Summary statistics aligned well with these visual trends.
# Following the distribution analysis, correlation analysis revealed strong positive correlation (visualisation 2) between ozone and temperature (r = 0.69), a moderate positive correlation with solar radiation (r = 0.34), and a moderate negative correlation with wind (r = -0.60). These findings guided further analysis.
# Regression Analysis- A regression model assessed the influence of solar radiation, wind, and temperature on ozone levels. Results indicated that a 1°F rise in temperature increases ozone by 1.66 ppb, and a unit rise in solar radiation by 0.06 ppb. These effects, though modest, were statistically significant (p < 0.05). Multicollinearity tests confirmed low inter-correlation among predictors, supporting model reliability. Solar radiation had minimal ozone-depleting influence, aligning with the historical context of CFC-related impacts.
# Scatterplots (visualisation 3) supported earlier findings: ozone and temperature exhibited a positive linear trend; ozone and solar radiation showed a weak-to-moderate relationship—disturbed by outliers in August; ozone and wind lacked a clear pattern. These visuals validated the observed regression outcomes. Temporal analysis (visualisation 4) revealed irregular ozone fluctuations, peaking mid-to-late August—consistent with summer conditions. The relationship with solar radiation weakened over time, generally declining as autumn approached, aside from a few anomalies.

# Limitations- Despite its value, the dataset presents limitations. Observations span diverse sites with distinct characteristics—urban, natural, and insular—each with its own microclimate. Moreover, data were collected at varying times, introducing spatial and temporal heterogeneity that may confound results. These inconsistencies highlight the need for standardized, temporally and spatially aligned data collection. Nonetheless, this analysis provides a foundation for further exploration of seasonal and anthropogenic impacts on air quality.

```{r plots, echo=FALSE}

# Boxplots
# Plot boxplots for each variable grouped by Month
boxplot <- ggplot(ozone_long, aes(x = factor(Month), y = Value)) +
  geom_boxplot() +
  facet_wrap(~ Variable, scales = "free_y") +  # separate plot for each variable, free y-axis scale
  labs(x = "Month", y = "Value", title = "Monthly Boxplots of Variables") +
  theme_minimal()

boxplot_jitter <- ggplot(ozone_long, aes(x = factor(Month), y = Value)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # add daily points with some horizontal jitter
  facet_wrap(~ Variable, scales = "free_y") +
  labs(x = "Month", y = "Value", title = "Monthly Boxplots with Daily Points") +
  theme_minimal()

print(boxplot)
print(boxplot_jitter)

ggsave("../figures/re_airquality_boxplot_visualisation1.jpg", plot = boxplot, width = 8, height = 6, dpi = 300)

ggsave("../figures/re_airquality_boxplot1_jitter_supplement.jpg", plot = boxplot_jitter, width = 8, height = 6, dpi = 300)

# Correlation Matrix
# correlation matrix_visualization2
corr_matrix <- cor(ozone_clean)

corrplot(corr_matrix, method = "circle", type = "lower", tl.cex = 0.8)

print(corrplot)


# Time series plots
# adding date from raw data 
ozone_raw <- ozone %>%
  mutate(Date = make_date(year = 1973, month = Month, day = Day))

# Create a copy of ozone_clean
ozone_clean_new <- ozone_clean

# Add the Date column from ozone_raw, but only up to the number of rows in ozone_clean
ozone_clean_new$Date <- ozone_raw$Date[1:nrow(ozone_clean_new)]

# Ozone vs Solar.R
temporal_ozone_solar <- ggplot(data = ozone_clean_new, aes(x = Date, y = Ozone)) +
  geom_line() +
  geom_point(aes(color = Solar.R), size = 1) +
  labs(
    title = "Ozone Levels Over Time (colored by Solar.R)",
    x = "Date",
    y = "Ozone (ppb)",
    color = "Solar.R (lang)"
  ) +
  scale_color_viridis_c(option = "D", direction = -1) +
  theme_classic()

print(temporal_ozone_solar)

ggsave("../figures/re_airquality_temporal_ozone_solar_visualisation4_1.jpg", plot = temporal_ozone_solar, width = 8, height = 6, dpi = 300)


# Ozone vs Wind
temporal_ozone_wind <- ggplot(data = ozone_clean_new, aes(x = Date, y = Ozone)) +
  geom_line() +
  geom_point(aes(color = Wind), size = 1) +
  labs(
    title = "Ozone Levels Over Time (colored by Wind)",
    x = "Date",
    y = "Ozone (ppb)",
    color = "Wind (mph)"
  ) +
  scale_color_viridis_c(option = "D", direction = -1) +
  theme_classic()

print(temporal_ozone_wind)

ggsave("../figures/re_airquality_temporal_ozone_wind_visualisation4_3.jpg", plot = temporal_ozone_wind, width = 8, height = 6, dpi = 300)

# Ozone vs Temp
temporal_ozone_temp <- ggplot(data = ozone_clean_new, aes(x = Date, y = Ozone)) +
  geom_line() +
  geom_point(aes(color = Temp), size = 1) +
  labs(
    title = "Ozone Levels Over Time (colored by Temp)",
    x = "Date",
    y = "Ozone (ppb)",
    color = "Temp (°F)"
  ) +
  scale_color_viridis_c(option = "D", direction = -1) +
  theme_classic()

print(temporal_ozone_temp)

ggsave("../figures/re_airquality_temporal_ozone_temp_visualisation4_2.jpg", plot = temporal_ozone_temp, width = 8, height = 6, dpi = 300)


# cluster plots
ozone_clean_month <- ozone_raw
ozone_clean_month$month_name <- month(ozone_raw$Month, label = TRUE)

# ozone vs solar.r
cluster_ozone_solar <- ggplot(
  data = ozone_clean_month,
  aes(x = Solar.R, y = Ozone)
) +
  geom_point(alpha = 0.4) +
  geom_smooth(formula = y ~ x + 0, method = "lm", color = "red", se = FALSE) +
  labs(
    x = expression(paste("Solar.R (lang)")),
    y = expression(paste("Ozone (ppb)")),
    caption = "Scatter plot of Ozone vs Solar Radiation for each month from May to September. 
               Red lines show linear regression fits with zero intercept."
  ) +
  facet_wrap(~month_name) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, margin = margin(t = 10, b = 10), size = 9),
    plot.caption.position = "plot"
  )

print(cluster_ozone_solar)

ggsave("../figures/re_airquality_cluster_ozone_solar_visualisation3_1.jpg", plot = cluster_ozone_solar, width = 8, height = 6, dpi = 300)


# ozone vs wind
cluster_ozone_wind <- ggplot(
  data = ozone_clean_month,
  aes(x = Wind, y = Ozone)
) +
  geom_point(alpha = 0.4) +
  geom_smooth(formula = y ~ x + 0, method = "lm", color = "red", se = FALSE) +
  labs(
    x = expression(paste("Wind (mph)")),
    y = expression(paste("Ozone (ppb)")),
    caption = "Scatter plot of Ozone vs Wind for each month from May to September. 
               Red lines show linear regression fits with zero intercept."
  ) +
  facet_wrap(~month_name) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, margin = margin(t = 10, b = 10), size = 9),
    plot.caption.position = "plot"
  )

print(cluster_ozone_wind)

ggsave("../figures/re_airquality_cluster_ozone_wind_visualisation3_3.jpg", plot = cluster_ozone_wind, width = 8, height = 6, dpi = 300)


# ozone vs Temp
cluster_ozone_temp <- ggplot(
  data = ozone_clean_month,
  aes(x = Temp, y = Ozone)
) +
  geom_point(alpha = 0.4) +
  geom_smooth(formula = y ~ x + 0, method = "lm", color = "red", se = FALSE) +
  labs(
    x = expression(paste("Temp (°F)")),
    y = expression(paste("Ozone (ppb)")),
    caption = "Scatter plot of Ozone vs Temperature for each month from May to September. 
               Red lines show linear regression fits with zero intercept."
  ) +
  facet_wrap(~month_name) +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0, margin = margin(t = 10, b = 10), size = 9),
    plot.caption.position = "plot"
  )

print(cluster_ozone_temp)

ggsave("../figures/re_airquality_cluster_ozone_temp_visualisation3_2.jpg", plot = cluster_ozone_temp, width = 8, height = 6, dpi = 300)


```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
